### 2014-11-12 tex/heap.tex
updates to heap.tex

\subsection{Heaps as Bipartite Graphs}
An important aspect, and a novel contribution, of a symbolic heap is its
representation as a labeled bipartite graph consisting of
references, $\cfgnt{r}$, and constraint location pairs,
$\lp\phi\ \cfgnt{l}\rp$. The heap structure itself captures that the actual
location pointed to by a reference is conditional on aliasing
relationships within the heap. In this way, the graph represents sets
of heaps with the conditions under which those heaps exist.

The graph is expressed in in $\cfgnt{L}$, the location map, and
$\cfgnt{R}$, the reference map. As done with the environment,
$\cfgnt{L}$ and $\cfgnt{R}$ are treated as partial functions where
$\cfgnt{L}(r) = \{(\phi\ \cfgnt{l})\ ...\}$ is the set of
location-constraint pairs in the heap associated with the given
reference, and $\cfgnt{R}(\cfgnt{l},\cfgnt{f}) = \cfgnt{r}$ is the
reference associated with the given location-field pair in the
heap. Predicate calculus is used to describe the more complex updates
to the heap in the semantics.

There are several properties of the heap that are invariant and
important to the correctness.
\begin{compactitem}
\item \textbf{Immutable references}: a reference does not change in
  the location map; it always points to the same set of constraint
  location pairs.
\item \textbf{Reference partitions}: new references are drawn from
  three distinct partitions $\mathrm{init}_r()$, $\mathrm{fresh}_r()$,
  and $\mathrm{stack}_r()$. The references strictly increase in value
  on each call, and modular arithmetic is used to determine the
  partition to which a reference belongs.
\item \textbf{Determinism}: a reference in  $(\cfgnt{L}\ \cfgnt{R})$ cannot point to two locations at the same time.
\[
\begin{array}{l}
\forall \cfgnt{r} \in \cfgnt{L}^\leftarrow\ (\forall (\phi\ \cfgnt{l}),(\phi^\prime\ \cfgnt{l}^\prime) \in \cfgnt{L}(r)\ (\\
\ \ \ \ (\cfgnt{l} \neq \cfgnt{l}^\prime \vee \phi \neq \phi^\prime) \Rightarrow (\phi \wedge \phi^\prime = \cfgt{false}))
\end{array}
\]
$\cfgnt{L}^\leftarrow$ is the pre-image of the location map.
\item \textbf{Type consistent}: every location associated to a reference has the same type.
\[
\begin{array}{l}
\forall \cfgnt{r} \in \cfgnt{L}^\leftarrow\ (\forall (\phi\ \cfgnt{l}),(\phi^\prime\ \cfgnt{l}^\prime) \in \cfgnt{L}(r)\ (\\
\ \ \ \ (\mathrm{Type}\lp\cfgnt{l}\rp = \mathrm{Type}\lp\cfgnt{l}^\prime\rp)))
\end{array}
\]
\item \textbf{Null and Uninitialized}: The location $\cfgnt{l}_\mathit{null}$ is a special location in the
heap to represent null. It has a companion reference
$\cfgnt{r}_\mathit{null}$. The initial heap for the machine is defined
such that $\cfgnt{L}(\cfgnt{r}_\mathit{null}) =
\{(\cfgt{true}\ \cfgnt{l}_\mathit{null})\}$
\end{compactitem}

DEFINE $\rsym$ and $\rgse$.

>>>>

\subsection{Heaps as Bipartite Graphs}
\sjp{General comment: we are missing motivation for why we discuss both
the gse and the summary heap machines - perhaps this goes in the intro?
And the fact that the GSE machine implements lazy versus lazier, etc. and
why this decision}
An important part and novel contribution of summary heaps for
symbolic execution is the labeled bipartite graph representation
of the heap which enables a summary heap to represent an equivalence class of
concrete heaps--all of the concrete heaps that follow the same control
flow to a given point of execution. The equivalence class splits depending 
on control flow branches, partitioning the heaps depending on the 
produced outcome. \sjp{Last sentence should be clarified.}

The bipartite graph consists of 
references, $\cfgnt{r}$, and constraint-location pairs,
$\lp\phi\ \cfgnt{l}\rp$, capturing the fact that the
actual location pointed to by a reference is conditional on
aliasing relationships within the heap. In this way, the graph
represents sets of heaps with the conditions under which those heaps
exist. This representation separates the constraint
problem from the morphology problem, i.e, separates the \sjp{??}
from the shape of the heap, and also makes 
it unnecessary to rewrite the
path condition every time the heap is updated. Without these two
features,
%properties, 
it would not be possible to lazily initialize heap
locations along a program path.

The graph is expressed in the location map, $\cfgnt{L}$, and
the reference map, $\cfgnt{R}$. As done with the environment,
$\cfgnt{L}$ and $\cfgnt{R}$ are treated as partial functions where
$\cfgnt{L}(r) = \{(\phi\ \cfgnt{l})\ ...\}$ is the set of
constraint-location pairs in the heap associated with the given
reference, and $\cfgnt{R}(\cfgnt{l},\cfgnt{f}) = \cfgnt{r}$ is the
reference associated with the given location-field pair in the
heap. Predicate calculus is used to describe the more complex updates
to the heap in the semantics.

There are several properties of the summary heap 
enforced by the rewrite rules that are invariant and
important to the correctness proofs
presented in \sjp{need sec ref}.
\begin{compactitem}
\item \textbf{Immutable references}: a reference does not change in
  the location map; it always points to the same set of constraint
  location pairs. This property ensures that constraints never need to
  be rewritten as the program evolves its state.
\item \textbf{Reference partitions}: new references are drawn from
  three distinct partitions (i) $\mathrm{init}_r()$ for initialization, (ii)
  $\mathrm{fresh}_r()$ for the input heap, and (iii) $\mathrm{stack}_r()$
  for references in the expressions and continuations. The references
  strictly increase in value on each call, and modular arithmetic is
  used to determine the partition to which a reference belongs. Reference
  partitions are a critical piece to the existence proof of a
  bisimulation that relates states with a concrete heap to states with
  a summary heap (see \sjp{add secref}).
\item \textbf{Determinism}: a reference in  $(\cfgnt{L}\ \cfgnt{R})$ cannot point to two  locations at the same time \sjp{multiple locations simultaneously?}.
\[
\begin{array}{l}
\forall \cfgnt{r} \in \cfgnt{L}^\leftarrow\ (\forall (\phi\ \cfgnt{l}),(\phi^\prime\ \cfgnt{l}^\prime) \in \cfgnt{L}(r)\ (\\
\ \ \ \ (\cfgnt{l} \neq \cfgnt{l}^\prime \vee \phi \neq \phi^\prime) \Rightarrow (\phi \wedge \phi^\prime = \cfgt{false}))
\end{array}
\]
$\cfgnt{L}^\leftarrow$ is the pre-image of the location map. \sjp{need more of a description here}
\item \textbf{Type consistency}: all locations associated with a reference have the same type.
\[
\begin{array}{l}
\forall \cfgnt{r} \in \cfgnt{L}^\leftarrow\ (\forall (\phi\ \cfgnt{l}),(\phi^\prime\ \cfgnt{l}^\prime) \in \cfgnt{L}(r)\ (\\
\ \ \ \ (\mathrm{Type}\lp\cfgnt{l}\rp = \mathrm{Type}\lp\cfgnt{l}^\prime\rp)))
\end{array}
\]
\item \textbf{Null and Uninitialized}: every summary heap contains 
a special location for null ($\cfgnt{l}_\mathit{null}$), and a special
location for uninitialized
  ($\cfgnt{l}_\mathit{un}$), with corresponding references
  $\cfgnt{r}_\mathit{null}$ and
  $\cfgnt{r}_\mathit{un}$, respectively. $\cfgnt{L}(\cfgnt{r}_\mathit{null}) =
  \{(\cfgt{true}\ \cfgnt{l}_\mathit{null})\}$ and
  $\cfgnt{L}(\cfgnt{r}_\mathit{un}) =
  \{(\cfgt{true}\ \cfgnt{l}_\mathit{un})\}$.
\end{compactitem}

\subsection{State Transition Relation}

The initial state of a Javalite program is constructed from its
surface syntax on the $\cfgnt{P}$ production:
$\lp\mu\ \lp\cfgnt{C}\ \cfgnt{m}\rp\rp$. Assuming that
$\lp\cfgnt{L}\ \cfgnt{R}\rp$ is an empty heap with no references
or locations and $\eta$ is an empty environment with no
variables, the initial state is
$$
\begin{array}{l}
\lp\mu 
\ \cfgnt{L}[\cfgnt{r}_\mathit{null} \mapsto \{\lp\cfgt{true}\ \cfgnt{l}_\mathit{null})\}\rp] 
           [\cfgnt{r}_\mathit{un} \mapsto \{\lp\cfgt{true}\ \cfgnt{l}_\mathit{un}\rp\}] \\
\ \cfgnt{R}
\ \cfgt{true}\ \eta\  \lp\lp\cfgt{new}\ \cfgnt{C}\rp\ \cfgt{@}\ \cfgnt{m}\ \cfgt{true}\rp\rp\ \cfgt{end}\rp
\end{array}
$$
The first expression in the initial state is a call to a constructor
to create an instance of the class $\cfgnt{C}$. \sjp{I was under the
impression that Javalite does not have constructors -- that
It only initializes the new object's locations on the heap with their default
values.} Once an instance of
$\cfgnt{C}$ exists, the method $\cfgnt{m}$ of the class is
invoked.

In general, the semantics of Javalite are expressed as
rewrites on strings using pattern matching. Consider the rewrite rule
for the beginning of a  field access instruction \sjp{in the Javalite
machine?}:
$$
\mprset{flushleft}
	\inferrule[Field Access(eval)]{}{
      \lp \cfgnt{L}\ \cfgnt{R}\ \phi\ \eta\ \lp \cfgnt{e}\ \cfgt{\$}\ \cfgnt{f}\rp \ \cfgnt{k}\rp  \rcom 
      \lp \cfgnt{L}\ \cfgnt{R}\ \phi\ \eta\ \cfgnt{e}\ \lp \cfgt{*}\ \cfgt{\$}\ \cfgnt{f} \rightarrow \cfgnt{k}\rp \rp 
	}
$$
If the string representing the current state matches the left side, then it
creates the new string on the right. In this example, the new string
on the right is now evaluating the expression $\cfgnt{e}$ in the field
access, and it includes the continuation indicating that it still
needs to complete the actual field access once the expression is
evaluated.

Other more complex rules such as the one to create a new instance of a
class define constraints on the rewrites and more complex
transformations on the heap.
$$
\mprset{flushleft}
	\inferrule[New]{
      \cfgnt{r} = \mathrm{stack}_r\lp \rp\\
      l = \mathrm{fresh}_l\lp \cfgnt{C}\rp\\\\
      \cfgnt{R}^\prime = \cfgnt{R}[\forall \cfgnt{f} \in \mathit{fields}\lp \mathrm{C}\rp \ \lp \lp l\ \cfgnt{f}\rp  \mapsto \cfgnt{r}_\mathit{null} \rp ] \\\\
      \cfgnt{L}^\prime = \cfgnt{L}[\cfgnt{r} \mapsto \{\lp \cfgt{true}\ l\rp \}]
    }{
      \lp \cfgnt{L}\ \cfgnt{R}\ \phi\ \eta\ \lp \cfgt{new}\ \cfgnt{C}\rp \ \cfgnt{k}\rp  \rcom
      \lp \cfgnt{L}^\prime\ \cfgnt{R}^\prime\ \phi\ \eta\ \cfgnt{r}\ \cfgnt{k}\rp 
	}
$$
In this rule, when the string matches the new-expression, it is rewritten to use
a new heap location where all of the fields for the new object point to
$\cfgnt{r}_\mathit{null}$
and the location map points a new stack reference to that new object.

In the remainder of the paper, we refer to the the generalized
symbolic execution with lazy initialization machine by the relation
$\rgse$ or $p \rgse p^\prime$, and to the new summary heap 
machine by $\rsym$ or $q \rsym q^\prime$.
Due to space restrictions, the full operational semantics for generalized 
symbolic execution with
lazy initialization for Javalite are presented in the supplemental
document. We also  present the portion of the Javalite 
rewrite rules which are common to both machines in 
the supplementary document. In the next section, we present
the rewrite rules specific to the new summary heap algorithm for symbolic
execution.
